\section{Concluding remarks}\label{conclusion}
The goal of this thesis was to explore vertex-labeled distance oracle in planar directed
graphs. We have given two oracles in the static setting. One described in Section
\ref{oracle1} which is simpler than the
other, but still has its applications when the number of labels is low, i.e below
$\sqrt{n}$, or when the desired space allocation is $o(n^{3/2})$. The other oracle in
Section \ref{oracle2} beats
the space-query product of the trivial solution and it matches the best theoretical
trade-offs for distance oracles in planar graph that are not vertex-labeled. Even though
it does not work for all distributions of $L$, it arguably works for many realistic
cases. \\
We also gave two distance oracles in the dynamic setting. The first result in Section
\ref{oracle3} tried to match the
trade-offs given by the works based upon FR-Djikstra. Previous works have not been able
to reduce the trade-off of FR-Djikstra with more than a factor of $O(\lg^2\lg n)$, but
have been able to produce oracles that extend it with other operations with the same
update times as well as handling negative edge-lengths. We have shown it can be extended to vertex-labeled planar graphs. The
oracle in Section \ref{oracle4} supports the classic update in vertex-labeled graph - the
label change. We managed to do this in expected constant time, though without the
assumption on the distribution of $L$, the update time will suffer.

\subsection{Reflections and future works}
The obvious improvement would be remove the assumption on the distribution of $L$. I
tried different strategies. My first observation was that we wanted, given a query
$Q(u,\lambda)$, to find the nearest $\lambda$-labeled vertex to $u$. This boils down to
the nearest neighbor problem in planar graphs (known by many names). If we could be able to construct a data structure
of size $O(n^{3/2})$ that could answer this problem in $O(\lg n)$ time, the overall time
bounds would not be affected. However, it seemed that for a single label, this would
require linear space, which would add up to $O(n\ell)$ space for all labels. Most
landmark based methods did not seem to help either. A whole line of work is dedicated to
nearest neighbour search in general metrics
\cite{krauthgamer2004navigating}\cite{krauthgamer2005black}\cite{karger2002finding}\cite{plaxton1999accessing}\cite{beygelzimer2006cover}\cite{cole2006searching}.
However, all of them are approximate results and often work under certain conditions
(growth-restricted, low doubling dimension etc.). One might hope to uncover an exact NNS
structure or even one that can return the $k$ nearest neighbours for
$k=O(\text{polylog}(n))$ as this would be sufficient to fix the gap in the distance
oracle. \\
Besides mending current results, one could research if the results could extend to related
families of graphs. In particular, the $O(n)$-separable graphs, bounded genus graphs,
graphs with bounded treewidth.
