\section{Introduction}\label{introduction}
Efficient algorithms on networks are playing an increasing important role today, and many
networks have structural properties that can be exploited to allow for much faster
algorithms than those in general graphs. One such class of networks are planar graphs,
which has its importance in
areas such as road transportation, computer vision and many more. \\
As these networks can be quite large, one problem is constructing a compact \textit{distance
oracle}, which is a data structure that is able to either give exact or approximate
shortest paths between two nodes quickly while not taking up too much space. They were
introduced by Thorup and Zwick \cite{thorup2005approximate}. This problem has since been
extensively studied. However, a less explored variant of this problem
was introduced by Hermelin et al. \cite{hermelin2011distance}. In the variant, we seek a
compact distance oracle on \textit{vertex-labeled} graphs. For this scenario, each node
has a label $\lambda\in L$ (e.g. "gas
station" or "supermarket") and the oracles answer shortest paths $\delta(u,\lambda)$ between a
vertex $u$ and
the nearest vertex with the given label $\lambda$. This generalizes the normal distance oracle problem since when each node is
associated with a unique label, they are equivalent. \\
When analysing the performance of
distance oracles, we typically look at three things. The first is \textit{query} time.
This is the amount of time it takes to return the shortest path between two given
vertices or the shortest path between a vertex and the nearest $\lambda$-labeled vertex.
The second is the \textit{space} required. This is the information stored in the data
structure we have access to at query time. The last is \textit{preprocessing}. This is
the time it takes to build the data structure. \\
One might also study the same problem in a dynamic setting, where the graph is prone to
other operations such as edge weight changes, edge insertion and/or deletions. Another
update specific to vertex-labeled graphs is the update of a label. As motivation,
consider the cases when new roads are built, torn down or when a library
becomes a restaurant. In this case, we also measure the performance in \textit{update}
time, which is the time it takes to adjust the distance oracle to reflect the change.

\subsection{Thesis outline}
This thesis focuses on vertex-labeled distance oracles in directed planar graphs that can provide
exact distances between a vertex and a desired label $\lambda$. In Section \ref{exactPlanar} we
give two oracles for the static setting. The first oracle described in Section \ref{oracle1}
exploits the fact that
the number of labels $\ell$ is less than the number of vertices in the graph $n$. This
allows us to construct a more compact oracle with smaller query times. Specifically, we
can in $\tilde{O}(n\ell^{2/3})$ time construct an oracle of size $O(nl^{2/3})$ that answer queries in
$O(\ell^{1/3})$ time. The second oracle
in Section \ref{oracle2}, is a rather technical oracle, which achieves better query times, but has no dependence
on $\ell$. This oracle does not work for certain distributions of the label set $L$. It can
in $\tilde{O}(n^{3/2})$ time preprocess a planar graph into an oracle requiring $O(n^{3/2})$
space that can answer queries in $O(\text{polylog}(n))$ time. \\
In Section \ref{dynamicPlanar}, we consider the dynamic setting. In Section
\ref{oracle3}, we give an oracle based on the well-known FR-Djikstra algorithm, that can handle edge
weight updates, edge insertions and deletion in amortized $\tilde{O}(n^{2/3})$ time. It
is able to answer queries in $\tilde{O}(\min\{\text{size}(\label)n^{2/3}, n\})$ time, where
  $\text{size}(\lambda)$ refer to the number of vertices with label $\lambda$. In Section
  \ref{oracle4}, we extend the oracle from Section \ref{oracle2} to support label
changes in expected amortized $O(1)$ time. That is, we can change a label $\lambda_i$ to
$\lambda_j$ while maintaining correct shortest paths. \\
For each of the two oracles in the static setting, we give trade-offs between space and
query times.

\subsection{Notation}\label{notation}
For a graph $G=(V,E)$ with label set $L=\{\lambda_1, \lambda_2, \dots, \lambda_\ell\}$,
we define $|V|=n$, $|E|=m$ and $|L|=\ell$. \\
\\
A distance $\delta(v,\lambda)$ for a $v\in V$ and $\lambda\in L$ is the distance between
the node $v$ and the nearest $\lambda$-labeled node. Likewise, we use $\delta(u,v)$ to denote
the minimal distance between two nodes $u,v\in V$. When we write $d(u,\lambda)$ or
$d(u,v)$, we do not necessarily mean the shortest path, but just \textit{a} distance.
This is used when the measured distance is not in the entire graph $G$, but just the shortest
path in a subgraph of $G$. \\
\\
We use $Q(v,\lambda)$ or $Q(u,v)$ to denote the queries "What is $\delta(v,\lambda)$?" or
"What is $\delta(u,v)$?" respectively. \\
\\
For the sake of consistency, we will refer to general vertices as $u$ or $v$. Some
vertices have the property of being \textit{boundary vertices} (described in Section
\ref{rdiv2}) and will refer to them as $b$.
